---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
set.seed(1)
```

# leftcens

<!-- badges: start -->
[![R-CMD-check](https://github.com/fabregithub/leftcens/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/fabregithub/leftcens/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/fabregithub/leftcens/actions/workflows/pkgdown.yaml/badge.svg)](https://fabregithub.github.io/leftcens/)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

**leftcens** provides descriptive statistics and imputation for left- and
interval-censored measurement data — the kind produced by analytical and
environmental laboratories, where a result may fall below a reporting threshold
rather than being a clean number.

A measurement is reported in one of three tiers, and each is really an *interval*
the true value lies inside:

| Report                     | What is known               | Interval         |
|----------------------------|-----------------------------|------------------|
| Non-detect                 | below the detection limit   | `(0, MDL)`       |
| Detected, not quantified   | between MDL and LCMRL        | `(MDL, LCMRL)`   |
| Quantified                 | an actual measurement        | `(value, value)` |

One data representation (`cens_data`) underlies everything, so the interval
semantics are defined exactly once and shared by two capability modules:

- **Descriptive statistics** — non-parametric (Turnbull NPMLE) and
  semi-parametric (proportional-hazards / proportional-odds) summaries of
  interval-censored data.
- **Imputation** — Gibbs-sampling imputation that draws each censored cell from
  within its own interval, extended to accept *per-cell* bounds so non-detect and
  detected-not-quantified values in the same analyte are handled correctly.

Classic single-limit (Helsel-style) left censoring is treated as a degenerate
two-tier case of the same model, not a separate mode.

## Scope — what `leftcens` is (and isn't) for

`leftcens` is for **descriptive statistics and reconstructing the marginal
distribution** of censored measurements: per-analyte summaries, quantiles, and
imputed values (with Rubin's-rules uncertainty via `gsimp_mi()`) for reporting.

It is **not** a tool for **regression or association modelling** with a censored
variable. If a censored analyte enters a model as a predictor (e.g. an exposure)
or an outcome, imputing it here first and plugging it in can **bias the estimated
association** — valid inference needs the censored value imputed *congenially with
the analysis model* (conditioning on the outcome), which a stand-alone
distribution-reconstruction step cannot do. For that, impute the censored variable
**inside** the analysis: a joint model that conditions on the outcome (`brms` with
`mi()` + `cens()`), or a censored imputation method within your own multiple-
imputation loop. See the *"How much can you trust the imputation?"* vignette.

## Installation

leftcens is not yet on CRAN. Install the development version from GitHub:

``` r
# install.packages("remotes")
remotes::install_github("fabregithub/leftcens")
```

## Getting started

`leftcens` ships two example lab exports; `as_cens_list()` turns either into one
`cens_data` per analyte, which flows into the descriptive and imputation tools:

```{r taste}
library(leftcens)

# `groundwater` is bundled with the package (a wide ".left"/".right" lab export)
cl <- as_cens_list(groundwater)               # one cens_data per analyte

quantile(desc_np(cl), probs = c(0.5, 0.9))    # per-analyte quantiles (NA below the QL)

filled <- gsimp_impute(build_bounds(cl))      # impute the censored cells (copula: skew-robust default)
```

Descriptive summaries are table-ready via `as.data.frame()` (`gt`, `flextable`,
`knitr::kable()`, …) and broom-style `tidy()` / `glance()`.

The four vignettes (also on the [package website](https://fabregithub.github.io/leftcens/))
carry the detail:

- **Getting started** — the interval data model and the descriptive and
  imputation tools.
- **Preparing your own data** — from a wide lab export (interval *or* flagged) to
  results, using the bundled `groundwater` / `surfacewater` datasets.
- **Per-cell imputation bounds** — the Gibbs sampler and a recovery
  demonstration.
- **How much can you trust the imputation?** — practical guidance and the
  reliability limit (imputation stays trustworthy to roughly 50% non-detects; the
  full Monte Carlo study is in `validation/`).

## License and provenance

leftcens is released under the **MIT license**.

The imputation is an independent, clean-room implementation of the GSimp *method*
(Wei R. *et al.* 2018, *PLoS Comput. Biol.* 14(1):e1005973,
[doi:10.1371/journal.pcbi.1005973](https://doi.org/10.1371/journal.pcbi.1005973)),
written from the published algorithm rather than ported from the (CC BY-NC-SA)
GSimp source. Please cite the GSimp paper when you use the imputation.
