---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
set.seed(1)
```

# leftcens

<!-- badges: start -->
[![R-CMD-check](https://github.com/fabregithub/leftcens/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/fabregithub/leftcens/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

**leftcens** provides descriptive statistics and imputation for left- and
interval-censored measurement data — the kind produced by analytical and
environmental laboratories, where a result may fall below a reporting threshold
rather than being a clean number.

A measurement is reported in one of three tiers, and each is really an *interval*
the true value lies inside:

| Report                     | What is known               | Interval         |
|----------------------------|-----------------------------|------------------|
| Non-detect                 | below the detection limit   | `(0, MDL)`       |
| Detected, not quantified   | between MDL and LCMRL        | `(MDL, LCMRL)`   |
| Quantified                 | an actual measurement        | `(value, value)` |

One data representation (`cens_data`) underlies everything, so the interval
semantics are defined exactly once and shared by two capability modules:

- **Descriptive statistics** — non-parametric (Turnbull NPMLE) and
  semi-parametric (proportional-hazards / proportional-odds) summaries of
  interval-censored data.
- **Imputation** — Gibbs-sampling imputation that draws each censored cell from
  within its own interval, extended to accept *per-cell* bounds so non-detect and
  detected-not-quantified values in the same analyte are handled correctly.

Classic single-limit (Helsel-style) left censoring is treated as a degenerate
two-tier case of the same model, not a separate mode.

## Installation

leftcens is not yet on CRAN. Install the development version from GitHub:

``` r
# install.packages("remotes")
remotes::install_github("fabregithub/leftcens")
```

## Example

Build a `cens_data` object from per-observation bounds and summarise it
non-parametrically:

```{r example}
library(leftcens)

# an analyte with MDL = 0.5, LCMRL = 2.0
x <- as_interval_data(
  left  = c(0,   0,   0.5, 3.1, 4.2, 0,   1.9),
  right = c(0.5, 0.5, 2.0, 3.1, 4.2, 0.5, 2.0)
)
x

fit <- desc_np(x)
quantile(fit, probs = c(0.1, 0.5, 0.9))
```

Impute censored cells of a wide analyte matrix, drawing each value from within
its interval. Naively substituting the detection limit biases means upward, since
censoring removes the *low* values; the Gibbs sampler draws from within each
interval and recovers the truth more closely (the more so with more, and more
correlated, analytes):

```{r impute}
# correlated log-concentrations, censored below each analyte's 30th percentile
n <- 200; p <- 3
z_true <- matrix(rnorm(n * p), n, p) %*% chol(matrix(0.6, p, p) + diag(0.4, p))
conc <- exp(z_true)
lim <- apply(conc, 2, quantile, 0.30)

cds <- lapply(seq_len(p), function(j) {
  cj <- conc[, j]
  as_interval_data(
    left  = ifelse(cj < lim[j], 0,      cj),
    right = ifelse(cj < lim[j], lim[j], cj)
  )
})

bounds <- build_bounds(cds)
filled <- gsimp_impute(bounds, iters_all = 30)

naive <- bounds$data_wide
naive[bounds$to_impute] <- bounds$hi_mat[bounds$to_impute]

rbind(
  truth = colMeans(z_true),
  gsimp = colMeans(filled),
  naive = colMeans(naive)
)
```

## Learn more

- `vignette("getting-started", package = "leftcens")` — the data model and the
  descriptive-statistics tools.
- `vignette("imputation-bounds", package = "leftcens")` — the per-cell bounds
  and the Gibbs sampler, with a recovery demonstration.

## How much censoring can be imputed reliably?

A Monte Carlo validation study (in `validation/`) maps where imputation stays
trustworthy. In short: imputation is reliable — small bias **and** calibrated
95% multiple-imputation coverage — down to roughly **75–80% detection (≤ ~20–25%
non-detects)**. Beyond that, coverage fails before bias does; larger samples do
*not* help (they tighten intervals around a slightly-biased centre, so inference
gets *worse*); and inter-analyte correlation does not extend the range, because
each analyte's reliability is governed by its own detection rate. Right-skewed
data is more forgiving than symmetric log-scale data. See the
`imputation-bounds` vignette and `validation/README.md` for details.

## License and provenance

leftcens is released under the **MIT license**.

The imputation is an independent, clean-room implementation of the GSimp *method*
(Wei R. *et al.* 2018, *PLoS Comput. Biol.* 14(1):e1005973,
[doi:10.1371/journal.pcbi.1005973](https://doi.org/10.1371/journal.pcbi.1005973)),
written from the published algorithm rather than ported from the (CC BY-NC-SA)
GSimp source. Please cite the GSimp paper when you use the imputation.
