---
title: "Getting started with leftcens"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with leftcens}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
set.seed(1)
```

```{r setup}
library(leftcens)
```

## The problem: measurements you only know an interval for

Analytical and environmental measurement data rarely come as clean numbers. A
laboratory reports a concentration in one of three ways, depending on where the
true value sits relative to two reporting thresholds --- the **method detection
limit** (MDL) and the **lowest concentration minimum reporting level** (LCMRL):

| Report            | What is known                    | Interval        |
|-------------------|----------------------------------|-----------------|
| *Non-detect*      | value is below the MDL           | `(0, MDL)`      |
| *Detected, not quantified* | value is between MDL and LCMRL | `(MDL, LCMRL)`  |
| *Quantified*      | an actual measurement            | `(value, value)`|

Every observation is really an **interval** the true value lies inside (a
quantified measurement is just an interval of width zero). `leftcens` takes this
view seriously: one representation, `cens_data`, underlies both the
descriptive-statistics and the imputation tools.

## The example data

The package ships `groundwater`, a small trace-metal data set in the shape a lab
would export it: one row per sample, and a `.left` / `.right` bound pair for each
of five metals.

```{r data}
# 11 columns (sample_id + 5 metals x .left/.right); first 3 metals shown
knitr::kable(head(groundwater[, 1:7]))
```

`as_cens_list()` reads the whole wide table at once, building one `cens_data`
object per analyte:

```{r cens-list}
cl <- as_cens_list(groundwater)   # metals inferred from the column names
cl
```

Each element is a `cens_data` object: it stores the per-observation `left`/`right`
bounds and *derives* the tier of every row (so the two can never disagree). By
default bounds are kept on the natural-log scale (`log_transform = TRUE`), the
scale the NPMLE and imputation machinery work on.

```{r one-analyte}
cl$As
```

For a single analyte you can build one directly with `as_interval_data(left,
right)`. Your own data will not always arrive as `.left`/`.right` columns ---
`vignette("preparing-data")` covers the common lab-export formats (qualifier
flags, single detection limits, ...) and how to turn each into this list.

## Non-parametric descriptive statistics

`desc_np()` fits the Turnbull non-parametric maximum-likelihood estimate of the
distribution behind the intervals --- no distributional assumption, and
non-detects and detected-not-quantified cells accounted for correctly. Give it
the whole list and it summarises every analyte at once:

```{r desc-np}
np <- desc_np(cl)
np                                 # analyte x percentile matrix
```

Any quantile below an analyte's **quantitation limit** is returned as `NA` ---
there it lies in the censored region where the fit is extrapolating, so a point
value would mislead. Pass `ql = 0` for the raw NPMLE estimates. `plot()` overlays
the estimated cumulative distributions:

```{r desc-np-plot}
plot(np)
```

## Semi-parametric descriptive statistics

`desc_sp()` fits a semi-parametric proportional-hazards (`"ph"`) or
proportional-odds (`"po"`) model with a non-parametric baseline. With covariates
it estimates how each shifts the concentration distribution. Suppose the samples
come from two sites:

```{r desc-sp}
site <- factor(rep(c("North", "South"), length.out = nrow(groundwater)))
sp <- desc_sp(cl, covariates = site, model = "ph")
sp
```

The reported ratio is a hazard ratio (an odds ratio under `"po"`); pass
`bs_samples > 0` for standard errors and p-values.

## Formatting results as a table

Both descriptive summaries have an `as.data.frame()` method that returns a tidy
data frame (with an `analyte` column for a multi-analyte fit), so a summary drops
straight into any table package --- `knitr::kable()`, `gt`, `flextable`, `DT`:

```{r table}
knitr::kable(as.data.frame(np, probs = c(0.5, 0.9)), digits = 2,
             caption = "Median and 90th-percentile concentration per metal")
```

For the tidyverse modelling ecosystem (`gtsummary`, `modelsummary`), broom-style
`tidy()` and `glance()` are available too:

```{r tidy-glance}
glance(np)
```

## Imputation

The other half of the package fills the censored cells --- for example to feed a
downstream model that cannot take intervals --- by Gibbs sampling, drawing each
value from within its interval. Pass the same list to `build_bounds()` and
`gsimp_impute()`:

```{r impute}
bounds <- build_bounds(cl)
filled_log <- gsimp_impute(bounds)          # default skew-robust ("copula") model

# results are on the log scale; back-transform to original units
filled <- round(exp(filled_log), 2)
knitr::kable(head(data.frame(sample_id = groundwater$sample_id, filled)),
             caption = "Completed analyte matrix (original units)")
```

Every imputed value lies within its interval by construction; observed cells are
returned unchanged. Before trusting imputed values, check whether imputation is
reliable for a data set like yours:

```{r preflight, eval = FALSE}
preflight_from_data(cl, n_rep = 30, M = 20)
```

## Where to next

- `vignette("preparing-data")` --- get *your* data into `leftcens`, whatever
  format it arrives in.
- `vignette("imputation-bounds")` --- how the per-cell bounds and the Gibbs
  sampler work, with a recovery demonstration.
- `vignette("imputation-guidance")` --- how much censoring can be imputed
  reliably (and how to check for your design).
- `vignette("reporting-guidance")` --- how to present censored-data results
  honestly in a report or paper.
