---
title: "Getting started with leftcens"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with leftcens}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
set.seed(1)
```

```{r setup}
library(leftcens)
```

## The problem: measurements you only know an interval for

Analytical and environmental measurement data rarely come as clean numbers. A
laboratory reports a concentration in one of three ways, depending on where the
true value sits relative to two reporting thresholds --- the **method detection
limit** (MDL) and the **lowest concentration minimum reporting level** (LCMRL):

| Report            | What is known                    | Interval        |
|-------------------|----------------------------------|-----------------|
| *Non-detect*      | value is below the MDL           | `(0, MDL)`      |
| *Detected, not quantified* | value is between MDL and LCMRL | `(MDL, LCMRL)`  |
| *Quantified*      | an actual measurement            | `(value, value)`|

Every observation is really an **interval** that the true value lies inside. A
quantified measurement is just an interval of width zero. `leftcens` takes this
interval view seriously: one data representation, `cens_data`, underlies both the
descriptive-statistics and the imputation tools.

## Building a `cens_data` object

`as_interval_data()` takes the per-observation lower and upper bounds and
classifies each row for you. Suppose an analyte has MDL = 0.5 and LCMRL = 2.0:

```{r construct}
x <- as_interval_data(
  left  = c(0,   0,   0.5, 3.1, 4.2, 0,   1.9),
  right = c(0.5, 0.5, 2.0, 3.1, 4.2, 0.5, 2.0)
)
x
```

The category is *derived* from the bounds, never supplied separately, so the two
can never disagree. By default the bounds are stored on the natural-log scale
(`log_transform = TRUE`), which is the scale the NPMLE and imputation machinery
work on; a non-detect's lower bound of `0` becomes `-Inf` there, exactly the
left-censored bound those methods expect.

### Coming from single-limit (Helsel-style) data

Classic left-censored data has one detection limit per observation and no
"detected but not quantified" band. That is simply the two-tier collapse of the
same interval model, so there is a convenience constructor for it:

```{r single}
as_interval_data_from_single(
  value    = c(2.0, 5.1, 2.0, 8.3),
  censored = c(TRUE, FALSE, TRUE, FALSE)
)
```

A censored value is treated as `(0, value)` (below the limit); a detected value
as `(value, value)`.

## Non-parametric descriptive statistics

`desc_np()` fits a non-parametric maximum-likelihood estimate (the Turnbull
NPMLE) of the distribution behind the intervals --- no distributional assumption,
and non-detects and detected-not-quantified cells are accounted for correctly.

```{r desc-np}
fit <- desc_np(x)
fit
```

Extract quantiles (on the original concentration scale) or plot the estimated
cumulative distribution:

```{r desc-np-quantile}
quantile(fit, probs = c(0.1, 0.5, 0.9))
```

```{r desc-np-plot}
plot(fit)
```

Two NPMLE algorithms are available: `method = "turnbull"` (the default, via
`survival::survfit()`) and `method = "wang"` (via `icenReg::ic_np()`). Because
the Turnbull NPMLE depends only on the ordering of the intervals, the estimate
is the same whether the data were stored on the log scale or not --- the fit is
computed on the original scale so quantiles read directly as concentrations.

## Semi-parametric descriptive statistics

`desc_sp()` fits a semi-parametric proportional-hazards (`"ph"`) or
proportional-odds (`"po"`) model with a non-parametric baseline. With no
covariates it estimates the baseline distribution; with covariates it estimates
how each covariate shifts the concentration distribution.

```{r desc-sp}
grp <- factor(c("A", "A", "A", "B", "B", "B", "B"))
sp <- desc_sp(x, covariates = grp, model = "ph")
sp
```

The reported ratio is a hazard ratio under `"ph"` (or an odds ratio under
`"po"`). Standard errors and p-values require bootstrap samples:

```{r desc-sp-bs}
sp_bs <- desc_sp(x, covariates = grp, model = "ph", bs_samples = 100)
summary(sp_bs)
```

You can overlay the estimated distribution for each covariate group:

```{r desc-sp-plot}
nd <- data.frame(covariate = factor(c("A", "B")))
plot(sp, newdata = nd)
```

## Imputation

The second half of the package imputes the censored cells of a wide analyte
matrix by Gibbs sampling, respecting each cell's interval. That has its own
walkthrough --- see `vignette("imputation-bounds", package = "leftcens")`.

```{r impute-teaser}
a <- as_interval_data(c(0, 2, 5, 8, 3), c(1, 4, 5, 8, 3))
b <- as_interval_data(c(3, 0, 7, 2, 6), c(3, 1, 7, 4, 6))
bounds <- build_bounds(list(analyte_a = a, analyte_b = b))
bounds
```
