---
title: "How much can you trust the imputation?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How much can you trust the imputation?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

```{r setup}
library(leftcens)
```

This vignette gives practical guidance for `gsimp_impute()`: which model to use,
how to set the knobs, how much censoring you can impute through, and how to
sanity-check your own data. It summarises a Monte Carlo validation study that
ships with the package source (in `validation/`); the numbers below come from
that study and are directional (moderate replication) --- see
`validation/RESEARCH_PLAN.md` for the full design and the definitive-run recipe.

## The short version

- **Use the default model (`imp_model = "tobit"`).** It fits an
  interval-censored regression on observed *and* censored rows, so it is not
  fooled by the fact that censored values are the small ones.
- **You can reliably impute an analyte with up to ~50% non-detects.** Below 50%
  ND the median is still directly estimable, and in simulation the default model
  keeps the mean essentially unbiased and its intervals calibrated across that
  whole range.
- **Defaults are ample:** `iters_all = 10` sweeps and `M = 20` imputations.
- **Co-impute a moderate set of analytes** (roughly 10-50) for best precision,
  though the method is valid well outside that range.
- **Work on the log scale** (`log_transform = TRUE`, the default).

## The reliability target: ND < 50%

The natural line is the **non-detect fraction**. If fewer than half an analyte's
values are non-detects, its median sits at or above the detection limit and is
directly recoverable; that makes ND < 50% the sensible boundary for "worth
imputing." We call an analyte *reliably imputed* when the mean bias is small
(|bias| <= 0.10 on the log scale) **and** the 95% multiple-imputation interval
covers the truth ~95% of the time.

In the validation study the default model met both criteria up to ~49% ND, with
the median recovered exactly (it is not even censored below 50% ND). The
observed-only models (`ridge`, `lm`, `glmnet`) only reach ~25% ND before their
intervals stop covering --- which is why `tobit` is the default.

## What the validation established

Across a range of conditions, the default (`tobit`) held the ND < 50% target;
the observed-only models did not. Summary of the drivers in `validation/`:

| Question | Finding for `tobit` |
|---|---|
| Correlation strength | Rescue holds even at zero correlation --- it comes from the per-analyte censored fit, not borrowed information. |
| Number of analytes (`p`) | Holds from `p = 3` up to `p = 200` (`p/n = 2`, PCA regime). RMSE is lowest around `p = 12-50`. |
| Non-linear dependence | Rescue holds; unmodelled non-linear dependence costs *precision* (RMSE), not validity. |
| Detected-not-quantified band | Holds as the DNQ band widens; the median is recovered even when it falls inside the interval-censored band. |
| Convergence | Point estimates settle by `iters_all ~ 3-5`; interval width by `M ~ 10-20`. |

The one consistent cost is **efficiency, not bias**: when dependence is
non-linear the intervals are wider than they need to be (mildly conservative),
because the linear conditional model cannot exploit that dependence.

## Choosing the settings

**`imp_model`.** Keep the default `"tobit"` unless you have a specific reason:

- `"tobit"` --- interval-censored Gaussian; the reliable choice, several times
  slower than ridge. Reduces predictors by PCA when analytes outnumber samples.
- `"ridge"` --- fast, always solvable; use for a quick look or when you only
  need a rough fill and know censoring is light (< ~25% ND).
- `"glmnet"` --- elastic net, closest to the original GSimp; needs the package.
- a custom `function(yo, Xo, Xm)` returning `list(mean, sd)`.

**`iters_all` (Gibbs sweeps) and `M` (imputations).** `iters_all = 10` and
`M = 20` are ample at target-edge censoring; heavier regimes may warrant more.

**How many analytes to co-impute.** Impute related analytes together --- the
conditional model borrows strength across them. Precision was best in the
`p = 12-50` range; very few analytes give a weak model, very many (`p >> n`)
lose a little to the PCA reduction, but both remain valid.

## Sanity-check your own data

The target is per-analyte, so first check each analyte's non-detect fraction.
`cens_data` already classifies every observation, so this is one line:

```{r nd-fraction}
# a small example: three analytes with 30% / 45% / 20% non-detects
set.seed(1)
n <- 80
z <- matrix(rnorm(n * 3), n, 3) %*% chol(matrix(0.5, 3, 3) + diag(0.5, 3))
conc <- exp(sweep(z, 2, c(0, 0.5, 1.0), "+"))
nd_target <- c(0.30, 0.45, 0.20)
analytes <- lapply(seq_len(3), function(j) {
  lim <- stats::quantile(conc[, j], nd_target[j])
  as_interval_data(ifelse(conc[, j] < lim, 0, conc[, j]),
                   ifelse(conc[, j] < lim, lim, conc[, j]))
})

nd_fraction <- function(x) mean(x$category == "non_detect")
vapply(analytes, nd_fraction, numeric(1))
```

Analytes at or above 50% ND are outside the reliable range --- report those with
extra caution (or only their upper-bounded summaries) rather than trusting a
point imputation. Then impute the rest together:

```{r impute}
bounds <- build_bounds(analytes)
bounds

filled <- gsimp_impute(bounds, iters_all = 10)   # tobit by default

# every imputed value lands inside its interval, by construction
m <- bounds$to_impute
all(filled[m] <= bounds$hi_mat[m] &
      (is.infinite(bounds$lo_mat[m]) | filled[m] >= bounds$lo_mat[m]))
```

Imputed values are on the log scale (matching `log_transform = TRUE`);
exponentiate for original units. For a full assessment tailored to *your* study
design --- your `n`, analyte count, correlation, censoring pattern --- adapt the
drivers in `validation/`, which simulate a design and report the bias and
coverage you can expect before you trust imputation on the real data.

## Caveats

- **> 50% ND:** below the target line; the median itself is censored and point
  imputation is no longer dependable. Prefer interval-censored *descriptive*
  summaries (`desc_np()`, `desc_sp()`) over imputed point values.
- **Very small `n`:** with only a handful of observed values per analyte the
  conditional fit is unstable regardless of model.
- **Strongly non-linear dependence:** valid but less precise (wider intervals);
  a non-linear censored model is a planned extension.
- These figures are from a simulation study; validate against your own design
  before relying on imputation for inference.
