---
title: "How much can you trust the imputation?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How much can you trust the imputation?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

```{r setup}
library(leftcens)
```

This vignette gives practical guidance for `gsimp_impute()`: which model to use,
how to set the knobs, how much censoring you can impute through, and how to
sanity-check your own data. It summarises a Monte Carlo validation study that
ships with the package source (in `validation/`); the numbers below come from
that study --- now confirmed by a definitive high-replication run (500
replications, `M = 50`) --- see `validation/RESEARCH_PLAN.md` for the full design.

## The short version

- **Use the default model (`imp_model = "tobit"`).** It fits an
  interval-censored regression on observed *and* censored rows, so it is not
  fooled by the fact that censored values are the small ones.
- **The median is reliably recovered up to ~50% non-detects** --- whether the
  data are log-normal or right-skewed. It is the estimand to lead with: below 50%
  ND the median sits at or above the detection limit, so it is directly
  estimable.
- **The log-scale *mean* is equally reliable when the data are roughly
  log-normal**, but under strong right-skew it is biased downward once censoring
  passes ~35% ND (see below) --- report the median, and check with a pre-flight
  run when skew is visible.
- **Defaults are ample:** `iters_all = 10` sweeps and `M = 20` imputations.
- **Co-impute a moderate set of analytes** (roughly 10-50) for best precision,
  though the method is valid well outside that range.
- **Work on the log scale** (`log_transform = TRUE`, the default).

## The reliability target: ND < 50%

The natural line is the **non-detect fraction**. If fewer than half an analyte's
values are non-detects, its median sits at or above the detection limit and is
directly recoverable; that makes ND < 50% the sensible boundary for "worth
imputing." We call an analyte *reliably imputed* when the mean bias is small
(|bias| <= 0.10 on the log scale) **and** the 95% multiple-imputation interval
covers the truth ~95% of the time.

The definitive run (500 replications, `M = 50`) shows this target holds with one
important qualification --- **it depends on the shape of the margins, not just
the censoring fraction:**

- **The median is recovered across the whole ND < 50% range, log-normal or
  skewed.** Below 50% ND the median is at or above the limit, so it is observed
  rather than imputed; its bias was ~0 in every condition. This is the estimand
  to report.
- **The log-scale mean is reliable only when the data are approximately
  log-normal.** With symmetric margins the default keeps the mean unbiased and
  its intervals calibrated all the way to ~55% ND. But under substantial
  right-skew the mean acquires a downward bias that grows with censoring --- small
  at 15% ND, but exceeding the |bias| <= 0.10 tolerance by ~35% ND --- because the
  linear-Gaussian conditional model underestimates the skewed lower tail it is
  imputing into. With larger `n` the intervals also tighten around that bias, so
  coverage can fall *below* nominal (bias, not variance, then dominates).

A skew sweep pins down how much skew the mean tolerates before it crosses the
|bias| <= 0.10 line --- and it shrinks as non-detects rise (the bias scales with
skewness *times* the non-detect fraction, `bias ~ 0.85 * skew * ND`):

| Non-detect fraction | Max tolerable log-scale skewness (mean still within 0.10) |
|---|---|
| 15% | ~1.1 |
| 25% | ~0.8 |
| 35% | ~0.55 |
| 45% | ~0.45 |

Rule of thumb: the mean is dependable while (sample log-scale skewness) x (ND
fraction) is below ~0.18; above that, report the median instead. The median's
bias stayed at machine zero across the entire sweep.

The observed-only models (`ridge`, `lm`, `glmnet`) only reach ~25% ND before
their intervals stop covering even in the easy (symmetric) case --- which is why
`tobit` is the default.

## What the validation established

Across a range of conditions the default (`tobit`) recovered the median
throughout ND < 50%, and the mean too when the margins were roughly log-normal;
the observed-only models did neither. Summary of the drivers in `validation/`:

| Question | Finding for `tobit` |
|---|---|
| Marginal skew (distribution shape) | **The axis that governs mean recovery.** Symmetric/log-normal: mean unbiased to ~55% ND. Strong right-skew: mean biased downward, failing the tolerance by ~35% ND. The median stays exact either way. |
| Correlation strength | Rescue holds even at zero correlation --- it comes from the per-analyte censored fit, not borrowed information. |
| Number of analytes (`p`) | Holds from `p = 3` up to `p = 200` (`p/n = 2`, PCA regime). RMSE is lowest around `p = 12-50`. |
| Non-linear dependence | Rescue holds; unmodelled non-linear dependence costs *precision* (RMSE), not validity. |
| Detected-not-quantified band | Holds as the DNQ band widens; the median is recovered even when it falls inside the interval-censored band. |
| Convergence | Point estimates settle by `iters_all ~ 3-5`; interval width by `M ~ 10-20`. |

The one consistent cost is **efficiency, not bias**: when dependence is
non-linear the intervals are wider than they need to be (mildly conservative),
because the linear conditional model cannot exploit that dependence.

## Choosing the settings

**`imp_model`.** Keep the default `"tobit"` unless you have a specific reason:

- `"tobit"` --- interval-censored Gaussian; the reliable choice, several times
  slower than ridge. Reduces predictors by PCA when analytes outnumber samples.
- `"ridge"` --- fast, always solvable; use for a quick look or when you only
  need a rough fill and know censoring is light (< ~25% ND).
- `"glmnet"` --- elastic net, closest to the original GSimp; needs the package.
- a custom `function(yo, Xo, Xm)` returning `list(mean, sd)`.

**`iters_all` (Gibbs sweeps) and `M` (imputations).** `iters_all = 10` and
`M = 20` are ample at target-edge censoring; heavier regimes may warrant more.

**How many analytes to co-impute.** Impute related analytes together --- the
conditional model borrows strength across them. Precision was best in the
`p = 12-50` range; very few analytes give a weak model, very many (`p >> n`)
lose a little to the PCA reduction, but both remain valid.

## Sanity-check your own data

The target is per-analyte, so first check each analyte's non-detect fraction.
`cens_data` already classifies every observation, so this is one line:

```{r nd-fraction}
# a small example: three analytes with 30% / 45% / 20% non-detects
set.seed(1)
n <- 80
z <- matrix(rnorm(n * 3), n, 3) %*% chol(matrix(0.5, 3, 3) + diag(0.5, 3))
conc <- exp(sweep(z, 2, c(0, 0.5, 1.0), "+"))
nd_target <- c(0.30, 0.45, 0.20)
analytes <- lapply(seq_len(3), function(j) {
  lim <- stats::quantile(conc[, j], nd_target[j])
  as_interval_data(ifelse(conc[, j] < lim, 0, conc[, j]),
                   ifelse(conc[, j] < lim, lim, conc[, j]))
})

nd_fraction <- function(x) mean(x$category == "non_detect")
vapply(analytes, nd_fraction, numeric(1))
```

Analytes at or above 50% ND are outside the reliable range --- report those with
extra caution (or only their upper-bounded summaries) rather than trusting a
point imputation. Then impute the rest together:

```{r impute}
bounds <- build_bounds(analytes)
bounds

filled <- gsimp_impute(bounds, iters_all = 10)   # tobit by default

# every imputed value lands inside its interval, by construction
m <- bounds$to_impute
all(filled[m] <= bounds$hi_mat[m] &
      (is.infinite(bounds$lo_mat[m]) | filled[m] >= bounds$lo_mat[m]))
```

Imputed values are on the log scale (matching `log_transform = TRUE`);
exponentiate for original units.

### A pre-flight reliability check

For a fuller assessment tailored to *your* design, `preflight_from_data()`
reads the non-detect fractions from your `cens_data`, estimates the correlation,
marginals, and **marginal skewness** (the last by an interval-censored fit, so a
symmetric imputation cannot hide it), then simulates many datasets like yours,
imputes them, and reports the bias and 95% coverage you can expect --- flagging
which analytes are reliably imputable. Because it feeds the estimated skew into
the simulation, it will surface the skew-induced mean bias rather than assume
log-normality; when it detects material right-skew it points you to
`imp_model = "copula"`. (It is a Monte Carlo procedure, so it takes a little time;
increase `n_rep` / `M` for more precise estimates.)

```{r preflight, eval = FALSE}
preflight_from_data(analytes, n_rep = 30, M = 20)
#> <preflight> reliability for 3 analytes, n = 80 (tobit model)
#>   design: rho=0.49, dnq=0.00, skew=0.69 | 30 reps, M=20, iters=10
#>   reliable if |bias| <= 0.10 AND coverage >= 0.90 AND ND < 50%
#>
#>  analyte nd_frac mean_bias median_bias  rmse coverage reliable
#>        1    0.30    ...         ...      ...     ...      ...
#>        ...
#> Verdict: ... of 3 analytes reliably imputable.
#>   Right-skew detected (sinh-arcsinh delta ~ 0.69): the mean may be biased low.
#>   Consider imp_model = "copula" (skew-robust), or report the median.

# re-run with the skew-robust model if skew was flagged:
preflight_from_data(analytes, imp_model = "copula", n_rep = 30, M = 20)
```

Or describe a design directly with `preflight_reliability(n, nd_frac, ...)`
without any data in hand --- useful when planning a study. The reproducible
Monte Carlo study behind these recommendations lives in the source
`validation/` directory.

## Caveats

- **> 50% ND:** below the target line; the median itself is censored and point
  imputation is no longer dependable. Prefer interval-censored *descriptive*
  summaries (`desc_np()`, `desc_sp()`) over imputed point values.
- **Right-skewed margins:** the log-scale *mean* is biased downward once
  (skewness x ND) exceeds ~0.18 (see the threshold table above), even though the
  median stays exact. Lead with the median (or interval-censored descriptive
  summaries), and use a pre-flight run to gauge the mean's bias for your design.
- **Larger `n` does not fix a biased mean:** when the mean is biased (e.g. under
  skew), more samples tighten the intervals around the wrong center and *lower*
  coverage. Good coverage at small `n` is not by itself evidence of low bias.
- **Very small `n`:** with only a handful of observed values per analyte the
  conditional fit is unstable regardless of model.
- **Strongly non-linear dependence:** valid but less precise (wider intervals);
  a non-linear censored model is a planned extension.
- These figures are from a simulation study; validate against your own design
  before relying on imputation for inference.
