---
title: "Per-cell imputation bounds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Per-cell imputation bounds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
set.seed(2024)
```

```{r setup}
library(leftcens)
```

## From intervals to imputable cells

The descriptive tools (`vignette("getting-started")`) summarise censored data in
place. Sometimes you instead need *complete* values --- for example to feed a
downstream model that cannot take intervals. `leftcens` imputes the censored
cells of a wide analyte matrix (rows = samples, columns = analytes) by Gibbs
sampling, drawing each imputed value from within its own censoring interval.

Two pieces do the work:

1. `build_bounds()` turns one `cens_data` per analyte into the per-cell bound
   matrices the sampler needs.
2. `gsimp_impute()` runs the Gibbs sampler.

## Why *per-cell* bounds

The imputation lineage here is GSimp (Wei et al. 2018), which imputes
left-censored metabolomics data using a single detection limit **per analyte
column**: every missing cell in a column shares one upper bound.

Environmental data breaks that assumption. Within a single analyte you can have
both non-detects `(0, MDL)` *and* detected-but-not-quantified values
`(MDL, LCMRL)` --- two different bands, in the same column. So `leftcens`
generalises the truncation to a bound **per cell**: `lo_mat` and `hi_mat` carry
one `(lo, hi)` pair for every entry, not one per column. A non-detect and a
detected-not-quantified value in the same analyte are imputed within their
respective intervals.

`build_bounds()` assembles these matrices:

```{r build}
# two analytes, each mixing all three tiers
a <- as_interval_data(left  = c(0, 2, 5, 8, 0, 3),
                      right = c(1, 4, 5, 8, 1, 3))
b <- as_interval_data(left  = c(3, 0, 7, 0, 6, 2),
                      right = c(3, 1, 7, 1, 6, 4))
bounds <- build_bounds(list(analyte_a = a, analyte_b = b))
bounds
```

The result bundles four aligned matrices on the working (log) scale:

```{r inspect}
bounds$data_wide   # observed values; NA where a cell must be imputed
bounds$lo_mat      # lower bounds; NA where the cell is observed
bounds$hi_mat      # upper bounds
```

Because all three are built together, the missing pattern of `data_wide` is
*guaranteed* to match the bounds --- there is no way for the data and its
truncation limits to disagree. On the log scale a non-detect's `(0, MDL)` becomes
`(-Inf, log MDL)`, i.e. left-censoring, with no special-casing:

```{r left-cens}
identical(is.na(bounds$data_wide), bounds$to_impute)
```

Cells whose bounds are both `NA` (ordinary, unconstrained missing values) flow
through too, imputed from `(-Inf, Inf)`.

## Running the sampler

`gsimp_impute()` fills every censored cell. Each analyte is repeatedly redrawn
from its conditional distribution given the other analytes, with every draw
constrained to that cell's interval by a truncated normal:

```{r impute}
filled <- gsimp_impute(bounds, iters_all = 20)
filled
```

Every imputed value lands inside its interval by construction:

```{r within}
m <- bounds$to_impute
all(filled[m] <= bounds$hi_mat[m] &
      (is.infinite(bounds$lo_mat[m]) | filled[m] >= bounds$lo_mat[m]))
```

### Knobs

- **`imp_model`** --- the conditional model relating each analyte to the others.
  The default `"tobit"` is an interval-censored Gaussian regression fit on
  observed *and* censored rows, which avoids the selection bias of the
  observed-only models and keeps imputation reliable to much heavier censoring
  (it is several times slower, and reduces the predictor dimension by PCA when
  analytes outnumber samples). The observed-only alternatives are `"ridge"`
  (fastest, always solvable), `"lm"`, and `"glmnet"` (elastic net, closest to
  the original GSimp); you may also pass your own `function(yo, Xo, Xm)`
  returning `list(mean, sd)`.
- **`iters_all`** --- the number of outer Gibbs sweeps; this is the mixing /
  convergence control. Increase it until the imputed summaries stabilise.
- **`iters_each`** --- inner refinement draws per analyte per sweep. This only
  matters for *stochastic* conditional models (a cross-validated `glmnet` fit
  differs run to run); for the deterministic ridge default it is redundant,
  hence the default of `1`.
- **`initial`** --- `"bounds"` (draw from each analyte's observed mean/sd,
  truncated) or `"qrilc"` (needs the *imputeLCMD* package).

## Does it work? A recovery check

The honest test of an imputation method is whether it recovers known truth. We
simulate correlated log-concentrations, censor them into the three tiers at each
analyte's own limits, impute, and compare.

```{r recovery}
n <- 200; p <- 4
mu <- c(0.5, 1.0, 1.5, 0.8)
R <- matrix(0.6, p, p); diag(R) <- 1
z_true <- sweep(matrix(rnorm(n * p), n, p) %*% chol(R), 2, mu, "+")
conc <- exp(z_true)                       # original-scale ground truth

# censor: below the 25th percentile -> non-detect, 25-40th -> DNQ
mdl   <- apply(conc, 2, quantile, 0.25)
lcmrl <- apply(conc, 2, quantile, 0.40)
cds <- lapply(seq_len(p), function(j) {
  cj <- conc[, j]
  left  <- ifelse(cj < mdl[j], 0,      ifelse(cj < lcmrl[j], mdl[j],   cj))
  right <- ifelse(cj < mdl[j], mdl[j], ifelse(cj < lcmrl[j], lcmrl[j], cj))
  as_interval_data(left, right)
})
bounds <- build_bounds(cds)
mean(bounds$to_impute)                    # fraction of cells censored
```

```{r recovery-fit}
filled <- gsimp_impute(bounds, iters_all = 30)

# Compare against naive substitution at the detection limit.
naive <- bounds$data_wide
naive[bounds$to_impute] <- bounds$hi_mat[bounds$to_impute]

rbind(
  truth = colMeans(z_true),
  gsimp = colMeans(filled),
  naive = colMeans(naive)
)
```

Substituting the detection limit biases the analyte means upward (censoring
removes the *low* values). The Gibbs imputation, drawing from the conditional
distribution within each interval --- with the default censored (`"tobit"`)
model --- recovers the true means with little residual bias, and recovers the
between-analyte correlation structure as well:

```{r recovery-cor}
off <- upper.tri(diag(p))
c(gsimp = mean(abs((cor(filled) - cor(z_true))[off])),
  naive = mean(abs((cor(naive) - cor(z_true))[off])))
```

## How much censoring can be imputed reliably?

A Monte Carlo study accompanies the package (`validation/`, run separately from
the build) that maps where imputation stays trustworthy, using two criteria:
mean bias (log scale) within 0.10 **and** 95% multiple-imputation coverage of
the true mean at least 0.90.

The study first characterised the *observed-only* models (e.g. `"ridge"`): they
are reliable only to roughly **20--25% non-detects**, and beyond that fail in an
instructive way. **Coverage fails before bias does** (a small residual upward
bias soon exceeds the correctly-narrow MI standard error); **more samples make
inference worse** (the bias is systematic, so larger `n` only tightens intervals
around an off-centre estimate --- coverage at ~45% censoring fell from ~0.5 at
n = 100 to ~0.01 at n = 300); and **correlation between analytes does not help**,
because each analyte's model is fit on its own upper-truncated observed values.
The common cause is that selection bias, which is exactly what the censored
model removes.

The default `"tobit"` model, fit on observed **and** censored rows, extends the
reliable range substantially: in a paired comparison it holds bias near zero and
coverage near nominal down to at least **~55% non-detects**, at both small and
large `n`. Distribution shape still matters (right-skewed data is more
forgiving), and enough must remain observed to fit the conditional model, but in
practice the default keeps imputation trustworthy well into moderate-to-heavy
censoring. See `validation/README.md` for the full study and how to reproduce
it.

## Assumptions and caveats

- **Missingness is informative (MNAR).** Censoring depends on the value itself
  (small values fall below the limit). The truncated-normal draw models this
  directly; do not treat these as missing-at-random and impute with an unbounded
  method.
- **Work on the log scale.** Concentration data is typically log-normal and the
  conditional-normal model assumes approximate normality. Build your `cens_data`
  with `log_transform = TRUE` (the default). Imputed values are returned on that
  same log scale --- exponentiate if you need original units.
- **The conditional model matters.** Recovery depends on the other analytes being
  informative predictors. With few or uncorrelated analytes the imputation
  reverts toward the marginal interval, which is still valid but less precise.
- **Convergence is your responsibility.** Increase `iters_all` until imputed
  summaries stop drifting; the defaults are modest.

## Provenance and licensing

The Gibbs sampler is an independent, clean-room implementation of the GSimp
*method* (Wei et al. 2018, PLoS Comput. Biol. 14(1):e1005973), written from the
published algorithm rather than ported from the GSimp source code. This keeps
`leftcens` under its MIT license. Please cite the original GSimp paper when you
use the imputation.
