---
title: "How much can you trust the imputation?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How much can you trust the imputation?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

```{r setup}
library(leftcens)
```

This vignette gives practical guidance for `gsimp_impute()`: which model to use,
how many imputations to pool, how much censoring you can impute through, and how
to sanity-check your own data. It summarises a Monte Carlo validation study that
ships with the package source (in `validation/`); the numbers below come from
that study --- confirmed by two definitive high-replication runs (500
replications, `M = 50`) --- see `validation/PHASE3_SKEW_PLAN.md` for the design.

## The short version (best practice)

- **Lead with the median.** Below 50% non-detects the median sits at or above the
  detection limit, so it is recovered exactly whatever the distribution shape.
  Report it as your primary summary; treat analytes at >= 50% ND as out of range.
- **Keep the default skew-robust engine (`imp_model = "copula"`) for the mean.**
  Real measurement data are often more skewed than log-normal and you usually
  cannot diagnose the shape under censoring. The copula fits a flexible margin per
  analyte and keeps the mean near-unbiased across skew; on genuinely log-normal
  data it matches `tobit`. Switch to `tobit` (assumes log-normality) or `ridge`
  (fastest) only for known-log-normal data or very large/wide problems where speed
  dominates.
- **Always use *multiple* imputation, never a single completed dataset.** One
  imputation treats the filled values as known and under-covers. Generate several
  and pool with Rubin's rules --- `gsimp_mi()` does this and reports the fraction
  of missing information (FMI).
- **Choose the number of imputations by FMI, not sample size.** Let
  `gsimp_mi(adaptive = TRUE)` pick it, or use **`m` ~ 30 for the copula** (~20 for
  tobit); raise it toward `100 * FMI` under heavy censoring.
- **The other knobs are cheap:** `iters_all = 10` sweeps; co-impute a moderate set
  of related analytes (roughly 10-50); work on the log scale (the default).

## The reliability target: ND < 50%

The natural line is the **non-detect fraction**. If fewer than half an analyte's
values are non-detects, its median sits at or above the detection limit and is
directly recoverable; that makes ND < 50% the sensible boundary for "worth
imputing." We call an analyte *reliably imputed* when the mean bias is small
(|bias| <= 0.10 on the log scale) **and** the 95% multiple-imputation interval
covers the truth ~95% of the time.

The definitive run (500 replications, `M = 50`) shows this target holds with one
important qualification --- **it depends on the shape of the margins, not just
the censoring fraction:**

- **The median is recovered across the whole ND < 50% range, log-normal or
  skewed.** Below 50% ND the median is at or above the limit, so it is observed
  rather than imputed; its bias was ~0 in every condition. This is the estimand
  to report.
- **For the mean, the margin's *shape* is what matters --- so use the right
  engine.** With `tobit`, the mean is unbiased only when the data are roughly
  log-normal; under substantial right-skew its linear-Gaussian conditional
  underestimates the skewed lower tail, so the mean is biased downward once
  censoring is non-trivial (and with larger `n` the intervals tighten around that
  bias, *lowering* coverage). **`imp_model = "copula"` removes this** by fitting a
  flexible margin per analyte: in the capstone it held `|bias| <= 0.01` and
  near-nominal coverage across the whole skew range at ND < 50%, while matching
  `tobit` on log-normal data.

If you use `tobit`, a skew sweep pins down how much skew its mean tolerates before
it crosses the |bias| <= 0.10 line --- and it shrinks as non-detects rise (the
tobit bias scales as `~ 0.85 * skew * ND`):

| Non-detect fraction | Max log-scale skewness for tobit (mean within 0.10) |
|---|---|
| 15% | ~1.1 |
| 25% | ~0.8 |
| 35% | ~0.55 |
| 45% | ~0.45 |

Rule of thumb for `tobit`: its mean is dependable while (sample log-scale skewness)
x (ND fraction) stays below ~0.18. Above that, switch to `imp_model = "copula"` (no
such threshold) or report the median. The median's bias stayed at machine zero
across the entire sweep for both engines.

The observed-only models (`ridge`, `lm`, `glmnet`) only reach ~25% ND before their
intervals stop covering even in the easy (symmetric) case.

## What the validation established

Both engines recover the median throughout ND < 50%. For the mean, `tobit`
succeeds only on roughly log-normal margins, whereas `copula` succeeds across
skew; the observed-only models do neither. Summary of the drivers in
`validation/`:

| Question | Finding |
|---|---|
| Marginal skew (distribution shape) | **The axis that governs mean recovery.** `tobit`: mean unbiased to ~55% ND when symmetric, but biased downward under right-skew (fails by ~35% ND). `copula`: mean unbiased across skew (capstone \|bias\| <= 0.01 at ND < 50%, 48/48 vs tobit 36/48). Median stays exact for both. |
| Correlation strength | Rescue holds even at zero correlation --- it comes from the per-analyte censored fit, not borrowed information. |
| Number of analytes (`p`) | Holds from `p = 3` up to `p = 200` (`p/n = 2`, PCA regime). RMSE is lowest around `p = 12-50`. |
| Non-linear dependence | Rescue holds; unmodelled non-linear dependence costs *precision* (RMSE), not validity. |
| Detected-not-quantified band | Holds as the DNQ band widens; the median is recovered even when it falls inside the interval-censored band. |
| Convergence | Point estimates settle by `iters_all ~ 3-5`; interval width by `M ~ 10-20`. |

The one consistent cost is **efficiency, not bias**: when dependence is
non-linear the intervals are wider than they need to be (mildly conservative),
because the linear conditional model cannot exploit that dependence.

## Choosing the settings

**`imp_model`.** Keep the default `"copula"` unless you have a specific reason not to:

- `"copula"` --- **the default, robust choice.** Fits a flexible (sinh-arcsinh)
  margin per analyte, imputes in a latent-Gaussian space, and back-transforms;
  near-unbiased across skew and matches `tobit` on log-normal data. It is a
  *proper* multiple-imputation engine (it draws the margin parameters), so its
  intervals are honest --- at the cost of being the slowest option and wanting a
  few more imputations.
- `"tobit"` --- interval-censored Gaussian. Fast-ish and reliable *when the data
  are log-normal*, but biased under skew. Use it when you know the margins are
  symmetric, or for very large/wide problems.
- `"ridge"` --- fastest, always solvable; a quick look, or light censoring
  (< ~25% ND) on log-normal data.
- `"glmnet"` --- elastic net, closest to the original GSimp; needs the package.
- a custom `function(yo, Xo, Xm)` returning `list(mean, sd)`.

**`iters_all` (Gibbs sweeps).** `iters_all = 10` is ample; point estimates settle
by 3-5 sweeps.

**How many analytes to co-impute.** Impute related analytes together --- the
conditional model borrows strength across them. Precision was best in the
`p = 12-50` range; very few analytes give a weak model, very many (`p >> n`)
lose a little to the PCA reduction, but both remain valid.

## Multiple imputation: pooling and how many

A single imputed dataset treats the filled values as if they were observed, so its
intervals are too narrow --- it under-covers. Generate several imputations and pool
the per-analyte means with Rubin's rules; `gsimp_mi()` runs the loop, propagates
the imputation uncertainty into the standard errors, and reports the **fraction of
missing information (FMI)** that drives everything.

```{r mi, eval = FALSE}
fit <- gsimp_mi(bounds, imp_model = "copula", m = 30)
fit$pooled     # per-analyte estimate, CI, and fmi
```

**How many imputations?** The required `m` is set by the FMI (roughly the
non-detect fraction, adjusted for how well the observed data predict the missing
values), **not by the sample size** --- a large `n` does not let you use fewer.
Two practical routes:

- Let it choose: `gsimp_mi(x, imp_model = "copula", adaptive = TRUE)` adds
  imputations until the pooled interval stabilises.
- Or fix it: **`m` ~ 30 for `"copula"`** (~20 for `"tobit"`), and raise it toward
  `100 * FMI` when the reported FMI is high (heavy censoring). Because the copula
  is a proper engine it reports an honest FMI (~0.4 at 35% ND) and so wants more
  imputations than `tobit`, whose smaller `m` reflects that it under-propagates
  uncertainty --- not that it needs less.

**For a downstream analysis** (a regression, a Bayesian model): get the completed
datasets with `gsimp_mi(x, return_imputations = TRUE)`, fit your model on **each**
dataset, and pool the *analysis estimates* with Rubin's rules --- do not average
the imputed datasets and analyse once.

## Sanity-check your own data

The target is per-analyte, so first check each analyte's non-detect fraction.
`cens_data` already classifies every observation, so this is one line:

```{r nd-fraction}
# a small example: three analytes with 30% / 45% / 20% non-detects
set.seed(1)
n <- 80
z <- matrix(rnorm(n * 3), n, 3) %*% chol(matrix(0.5, 3, 3) + diag(0.5, 3))
conc <- exp(sweep(z, 2, c(0, 0.5, 1.0), "+"))
nd_target <- c(0.30, 0.45, 0.20)
analytes <- lapply(seq_len(3), function(j) {
  lim <- stats::quantile(conc[, j], nd_target[j])
  as_interval_data(ifelse(conc[, j] < lim, 0, conc[, j]),
                   ifelse(conc[, j] < lim, lim, conc[, j]))
})

nd_fraction <- function(x) mean(x$category == "non_detect")
vapply(analytes, nd_fraction, numeric(1))
```

Analytes at or above 50% ND are outside the reliable range --- report those with
extra caution (or only their upper-bounded summaries) rather than trusting a
point imputation. Then impute the rest together:

```{r impute}
bounds <- build_bounds(analytes)
bounds

filled <- gsimp_impute(bounds, imp_model = "copula", iters_all = 10)

# every imputed value lands inside its interval, by construction
m <- bounds$to_impute
all(filled[m] <= bounds$hi_mat[m] &
      (is.infinite(bounds$lo_mat[m]) | filled[m] >= bounds$lo_mat[m]))
```

This is one completed dataset; for inference, pool several with `gsimp_mi()`
(above). Imputed values are on the log scale (matching `log_transform = TRUE`);
exponentiate for original units.

### A pre-flight reliability check

For a fuller assessment tailored to *your* design, `preflight_from_data()`
reads the non-detect fractions from your `cens_data`, estimates the correlation,
marginals, and **marginal skewness** (the last by an interval-censored fit, so a
symmetric imputation cannot hide it), then simulates many datasets like yours,
imputes them (with the default `"copula"`), and reports the bias and 95% coverage
you can expect --- flagging which analytes are reliably imputable. Because it
feeds the estimated skew into the simulation, it surfaces the skew-induced mean
bias rather than assuming log-normality; if you switch to the faster
`imp_model = "tobit"` and it detects material right-skew, it points you back to
`"copula"`. (It is a Monte Carlo procedure, so it takes a little time; increase
`n_rep` / `M` for more precise estimates.)

```{r preflight, eval = FALSE}
preflight_from_data(analytes, n_rep = 30, M = 20)   # default: copula
#> <preflight> reliability for 3 analytes, n = 80 (copula model)
#>   design: rho=0.49, dnq=0.00, skew=0.69 | 30 reps, M=20, iters=10
#>   reliable if |bias| <= 0.10 AND coverage >= 0.90 AND ND < 50%
#>
#>  analyte nd_frac mean_bias median_bias  rmse coverage reliable
#>        1    0.30    ...         ...      ...     ...      ...
#>        ...
#> Verdict: ... of 3 analytes reliably imputable.

# the faster tobit model would flag the skew the default is protecting you from:
preflight_from_data(analytes, imp_model = "tobit", n_rep = 30, M = 20)
#>   ... Right-skew detected (sinh-arcsinh delta ~ 0.69): the mean may be biased low.
#>   Consider imp_model = "copula" (skew-robust), or report the median.
```

Or describe a design directly with `preflight_reliability(n, nd_frac, ...)`
without any data in hand --- useful when planning a study. The reproducible
Monte Carlo study behind these recommendations lives in the source
`validation/` directory.

## Caveats

- **> 50% ND:** below the target line; the median itself is censored and point
  imputation is no longer dependable. Prefer interval-censored *descriptive*
  summaries (`desc_np()`, `desc_sp()`) over imputed point values.
- **Single imputation under-covers:** always pool several with `gsimp_mi()`. A
  biased engine cannot be rescued by more imputations --- under skew, `tobit`'s
  coverage stays wrong at any `m`; the fix is the right engine (`"copula"`), not a
  larger `m`.
- **Right-skewed margins with `tobit`:** its mean is biased downward once
  (skewness x ND) exceeds ~0.18 (see the threshold table), even though the median
  stays exact. Switch to `imp_model = "copula"`, or report the median.
- **Larger `n` does not fix a biased mean:** when the mean is biased, more samples
  tighten the intervals around the wrong center and *lower* coverage. Good coverage
  at small `n` is not by itself evidence of low bias.
- **Very small `n`:** with only a handful of observed values per analyte the
  conditional fit (and the copula's margin fit) is unstable regardless of model.
- **Strongly non-linear dependence at heavy censoring:** the copula's coverage
  erodes a little (its latent conditional is linear); bias stays low. A non-linear
  censored conditional was tried and did not help --- see `validation/`.
- **Not for association modelling with a censored variable.** These tools
  reconstruct the *marginal* distribution — right for descriptive summaries. If a
  censored analyte is a **predictor or outcome in a regression**, imputing it here
  and plugging it in can bias the association: the imputation must be *congenial*
  with the analysis model (condition on the outcome). Impute it **inside** the
  model instead — a joint model (`brms` `mi()` + `cens()`) or a censored method
  within your own imputation loop.
- These figures are from a simulation study; validate against your own design
  before relying on imputation for inference.
