---
title: "Preparing your own data for leftcens"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing your own data for leftcens}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

```{r setup}
library(leftcens)
```

This vignette is the practical on-ramp: how to get a typical laboratory data set
into `leftcens`, run the descriptive summaries, and impute --- starting from a
wide table of per-analyte lower/upper bounds. If you have not met the data model
yet, skim `vignette("getting-started")` first.

## The one thing to get right

Everything in `leftcens` starts from **per-observation intervals**: for each
measurement you supply a lower bound `left` and an upper bound `right` that the
true value lies between. The three reporting tiers are just three interval
shapes:

| Tier | Meaning | `left` | `right` |
|---|---|---|---|
| Non-detect | below the detection limit (MDL) | `0` | `MDL` |
| Detected, not quantified | between MDL and the quantitation limit (LCMRL) | `MDL` | `LCMRL` |
| Quantified | an actual measurement | `value` | `value` |

So the whole task of "preparing your data" is: **produce a `left` and a `right`
column for each analyte.** Once you have those, `as_interval_data()` does the
rest (it even derives which tier each row is in).

## Starting from a wide lab export

A common export has one row per sample and a `left`/`right` pair per analyte ---
exactly the shape below (analytes `A`--`E`, each with `.left` / `.right`):

```{r make-data}
# --- a stand-in for your own data; replace this with read.csv("yourfile.csv") ---
analytes <- c("A", "B", "C", "D", "E")
mdl   <- c(A = 0.5, B = 1.0, C = 0.2, D = 5.0, E = 0.1)   # detection limits
lcmrl <- c(A = 2.0, B = 4.0, C = 1.0, D = 20,  E = 0.5)   # quantitation limits

n <- 10
truth <- sapply(analytes, function(a) rlnorm(n, meanlog = log(lcmrl[[a]]), sdlog = 1))
dat <- data.frame(ID = sprintf("S%02d", seq_len(n)))
for (a in analytes) {
  v <- truth[, a]
  dat[[paste0(a, ".left")]] <- round(
    ifelse(v < mdl[[a]], 0,        ifelse(v < lcmrl[[a]], mdl[[a]],   v)), 2)
  dat[[paste0(a, ".right")]] <- round(
    ifelse(v < mdl[[a]], mdl[[a]], ifelse(v < lcmrl[[a]], lcmrl[[a]], v)), 2)
}
```

```{r show-data}
knitr::kable(dat, caption = "Wide lab export: one row per sample, left/right per analyte")
```

Read your own file with `dat <- read.csv("yourfile.csv")` (or `readxl::read_excel()`)
so it has this shape. Notice how the tiers show up: a non-detect is `(0, MDL)`,
a detected-not-quantified value is `(MDL, LCMRL)`, and a quantified value has
`left == right`.

## Build one `cens_data` per analyte

Turn each analyte's `.left` / `.right` column pair into a `cens_data` object. A
short loop over the analyte names does it:

```{r build-cens}
cens_list <- lapply(analytes, function(a) {
  as_interval_data(left  = dat[[paste0(a, ".left")]],
                   right = dat[[paste0(a, ".right")]])
})
names(cens_list) <- analytes

cens_list$A
```

`as_interval_data()` classifies each row for you (the printout shows the tier
counts), and by default stores bounds on the log scale (`log_transform = TRUE`),
which is what the imputation and NPMLE machinery expect. Keep the `ID` column
aside --- `leftcens` works on the analyte columns; you re-attach IDs afterward by
row order.

## Descriptive summaries, per analyte

`desc_np()` gives a non-parametric distribution summary for one analyte. Loop
over the list to summarise all of them, and use the `as.data.frame()` method to
collect the quantiles into one tidy table:

```{r desc}
np <- lapply(cens_list, desc_np)

summary_tbl <- do.call(rbind, lapply(analytes, function(a) {
  cbind(analyte = a, as.data.frame(np[[a]], probs = c(0.25, 0.5, 0.75)))
}))
knitr::kable(summary_tbl, digits = 2,
             caption = "Quantiles per analyte (NA = below the quantitation limit)")
```

Quantiles that fall below an analyte's quantitation limit come back as `NA`:
there the estimate would be an extrapolation into the censored region rather than
a reliably quantified value. (Pass `ql = 0` to `as.data.frame()` /
`quantile()` if you want the raw estimates.)

## Impute the censored values

To fill the censored cells --- for example to feed a downstream model that cannot
take intervals --- pass the whole list to `build_bounds()` and `gsimp_impute()`:

```{r impute}
bounds <- build_bounds(cens_list)
filled_log <- gsimp_impute(bounds)          # default censored ("tobit") model

# results are on the log scale; back-transform to original units
filled <- round(exp(filled_log), 2)
filled_with_id <- data.frame(ID = dat$ID, filled)
knitr::kable(head(filled_with_id), caption = "Completed analyte matrix (original units)")
```

Every imputed value lies within its interval by construction; observed
(quantified) cells are returned unchanged.

### Will imputation be reliable for your data?

Before trusting imputed values, run a pre-flight check: it simulates data like
yours and reports the bias and coverage you can expect, flagging analytes with
too much censoring (see `vignette("imputation-guidance")`).

```{r preflight, eval = FALSE}
preflight_from_data(cens_list, n_rep = 30, M = 20)
```

## If your data looks different

You still just need a `left` and a `right` per measurement. Common starting
points:

**A value plus a "censored" flag (single detection limit).** This is the classic
Helsel-style format --- one number per row and a `TRUE`/`FALSE` for whether it was
below the limit:

```{r single}
value    <- c(2.0, 5.1, 0.8, 3.3)
censored <- c(FALSE, FALSE, TRUE, FALSE)   # TRUE = non-detect at that value
as_interval_data_from_single(value, censored)
```

**A value plus a per-row detection limit.** If non-detects are recorded as the
limit with a flag, build the pair yourself:

```{r perrow}
result <- c(2.0, NA, 5.1, NA)      # measured value, NA where not detected
dl     <- c(0.5, 0.5, 0.5, 0.7)    # detection limit for that row
nd     <- is.na(result)
left  <- ifelse(nd, 0,  result)
right <- ifelse(nd, dl, result)
as_interval_data(left, right)
```

**Fully quantified data (no censoring).** Set `left == right == value`; those
rows are treated as exact and nothing special happens.

## A short checklist

- One `left` and one `right` column per analyte, with `left <= right`.
- Non-detects: `left = 0`, `right = MDL`. Do **not** substitute the limit or
  `MDL/2` yourself --- let `leftcens` handle the censoring.
- Keep concentrations on their natural (original) scale; leave
  `log_transform = TRUE` (the default) so `left = 0` maps to the expected
  non-detect bound.
- Keep sample IDs in a separate column and re-attach by row order.
- Check each analyte's non-detect fraction (`mean(x$category == "non_detect")`)
  --- above ~50% ND, prefer interval-censored *descriptive* summaries over
  trusting imputed point values.
