---
title: "Preparing your own data for leftcens"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing your own data for leftcens}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

```{r setup}
library(leftcens)
```

This vignette is about one thing: **getting your data into `leftcens`**, whatever
shape it arrives in. Once it is in, `vignette("getting-started")` shows what to do
with it (descriptive summaries, imputation).

## The one thing to get right

Everything in `leftcens` starts from **per-observation intervals**: for each
measurement you supply a lower bound `left` and an upper bound `right` that the
true value lies between. The three reporting tiers are just three interval
shapes:

| Tier | Meaning | `left` | `right` |
|---|---|---|---|
| Non-detect | below the detection limit (MDL) | `0` | `MDL` |
| Detected, not quantified | between MDL and the quantitation limit (LCMRL) | `MDL` | `LCMRL` |
| Quantified | an actual measurement | `value` | `value` |

So the whole task is: **produce a `left` and a `right` for each measurement.**
The sections below cover the layouts labs actually export. `as_cens_list()`
handles the two wide layouts directly and returns a named list of `cens_data`
(one per analyte) --- exactly what `desc_np()`, `desc_sp()`, and `build_bounds()`
consume. Read your file first with `read.csv()` / `readxl::read_excel()`.

## Layout 1: lower/upper bound columns

Each analyte already has a `.left` and a `.right` column, carrying the full
three-tier structure. The bundled `groundwater` data set has this shape:

```{r lr-data}
# 11 columns (sample_id + 5 metals x .left/.right); first 3 metals shown
knitr::kable(head(groundwater[, 1:7]))
```

`as_cens_list()` infers the analytes from the `.left`/`.right` column names:

```{r lr-list}
cens_list <- as_cens_list(groundwater)
cens_list
```

Custom suffixes are supported (e.g. `left_suffix = "_lo"`), and you can name the
analytes explicitly with `analytes = c("As", "Cd", ...)`.

## Layout 2: a value plus a qualifier flag

More often each analyte has a **value column and a qualifier-flag column** --- the
value plus a code such as `"<"` / `"ND"` / `"U"` for a non-detect (where the
reported number is the detection or reporting limit) and a blank for a quantified
result. The bundled `surfacewater` data set has this shape:

```{r flag-data}
# 11 columns (sample_id + 5 metals x value/.flag); first 3 metals shown
knitr::kable(head(surfacewater[, 1:7]))
```

Use `format = "flagged"`, setting `nd_flags` to whichever codes mark a
non-detect. A flagged non-detect becomes `(0, value)` (the value being the
limit); anything else is an exact `(value, value)`:

```{r flag-list}
sw_cens <- as_cens_list(surfacewater, format = "flagged", nd_flags = c("<", "ND", "U"))
sw_cens
```

`as_cens_list()` ignores columns that are not analyte pairs (such as
`sample_id`), which you re-attach by row order afterward.

## Layout 3: a value plus a "censored" flag (single detection limit)

The classic Helsel-style format --- one number per row and a `TRUE`/`FALSE` for
whether it was below the limit --- has a dedicated constructor:

```{r single}
value    <- c(2.0, 5.1, 0.8, 3.3)
censored <- c(FALSE, FALSE, TRUE, FALSE)   # TRUE = non-detect at that value
as_interval_data_from_single(value, censored)
```

## Layout 4: a value plus a per-row detection limit

If non-detects are recorded separately from the limit, build the `left`/`right`
pair yourself and call `as_interval_data()`:

```{r perrow}
result <- c(2.0, NA, 5.1, NA)      # measured value, NA where not detected
dl     <- c(0.5, 0.5, 0.5, 0.7)    # detection limit for that row
nd     <- is.na(result)
as_interval_data(left  = ifelse(nd, 0,  result),
                 right = ifelse(nd, dl, result))
```

## Building the three-tier structure yourself

`format = "flagged"` produces the two-tier structure (non-detect vs quantified).
If you also have a detected-not-quantified qualifier (e.g. `"J"`) and know each
analyte's MDL and LCMRL, place those rows in the middle tier by constructing the
bounds directly:

```{r three-tier}
value <- c(0.3, 1.2, 4.8, 0.1)
flag  <- c("<", "J",  "",  "<")     # "<" non-detect, "J" detected-not-quantified
mdl <- 0.5; lcmrl <- 2.0
left  <- ifelse(flag == "<", 0,   ifelse(flag == "J", mdl,   value))
right <- ifelse(flag == "<", mdl, ifelse(flag == "J", lcmrl, value))
as_interval_data(left, right)
```

The same idea works column-by-column across a wide table, then combine the
per-analyte results into a named list (as `as_cens_list()` does internally).

**Fully quantified data (no censoring).** Set `left == right == value`; those
rows are treated as exact.

## A short checklist

- One `left` and one `right` per measurement, with `left <= right`.
- Non-detects: `left = 0`, `right = MDL`. Do **not** substitute the limit or
  `MDL/2` yourself --- let `leftcens` model the censoring.
- Keep concentrations on their natural (original) scale; leave
  `log_transform = TRUE` (the default) so `left = 0` maps to the expected
  non-detect bound.
- Keep sample IDs in a separate column and re-attach by row order.
- Check each analyte's non-detect fraction (`mean(x$category == "non_detect")`)
  --- above ~50%, prefer interval-censored *descriptive* summaries over trusting
  imputed point values.

With the data prepared, head to `vignette("getting-started")` to summarise and
impute it.
