---
title: "Reporting censored-data results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reporting censored-data results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(leftcens)
```

This vignette is about **how to present** censored-measurement results in a report
or paper. It complements `vignette("imputation-guidance")`, which is about *when*
imputation is trustworthy; here we assume you have run the analysis and need to
write it up honestly.

## Non-detects: report the limit, not a substitute

The single most important rule: **do not replace non-detects with a made-up
number** (zero, the detection limit, or `MDL/2`) and then report ordinary
statistics on the result. That bias is exactly what `leftcens` exists to avoid.
Instead:

- Report a non-detect as **`< MDL`** (or as the interval), and state the method
  detection limit (MDL) and, where relevant, the quantitation limit (LCMRL).
- Report summary statistics computed by the censoring-aware tools
  (`desc_np()` / `desc_sp()`), not by substitution.
- If you imputed, say so explicitly (see below) --- imputed values are model
  outputs, not measurements.

## Which statistics to report, and when

Whether a statistic is even estimable depends on how much of the data is
censored, per analyte. A useful hierarchy:

- **Detection frequency** (the percentage detected) is *always* reportable and
  should be, whatever the censoring --- it is a fact about the data.
- **The median** is directly estimable when **fewer than 50% are non-detects**
  (the median then sits at or above the limit).
- **The mean and low quantiles** are trustworthy only under light censoring; the
  `desc_np()` NPMLE estimates them accounting for censoring, but they become
  unreliable as non-detects grow (see `vignette("imputation-guidance")`).

`leftcens` records the limits and returns quantiles below the quantitation limit
as `NA`, which is precisely what you want when tabulating: a blank/`< LCMRL`
rather than a spurious number. A reporting table falls out of `glance()` and
`quantile()`:

```{r report-table}
cl <- as_cens_list(groundwater)
np <- desc_np(cl)

g <- glance(np)
q <- quantile(np, probs = c(0.5, 0.9))       # NA below the quantitation limit

report <- data.frame(
  Analyte      = g$analyte,
  n            = g$nobs,
  `Detected %` = round(100 * (1 - g$n_non_detect / g$nobs)),
  MDL          = g$detection_limit,
  LCMRL        = g$quantitation_limit,
  Median       = q[g$analyte, "50%"],
  `90th pct`   = q[g$analyte, "90%"],
  check.names  = FALSE
)
knitr::kable(report, digits = 2,
             caption = "A reportable summary: detection frequency, limits, and estimable quantiles")
```

Read an `NA` median as "**< LCMRL**": more than half of those samples were below
the quantitation limit, so the median is not a quantified value. Report it that
way rather than forcing a number. A limit shown as `NA` (e.g. the Zn `MDL` here)
just means that tier did not occur in the data --- no sample was a Zn non-detect,
so no detection limit is derivable; report your laboratory's stated limit
instead.

## When an analyte is too censored for point summaries

If an analyte has more than roughly 50% non-detects, do **not** report an imputed
mean or median as if measured. Report instead:

- the **detection frequency** and the **MDL/LCMRL**;
- optionally an **interval-censored descriptive summary** (`desc_np()` gives the
  NPMLE of the distribution, and any estimable upper quantiles);
- a statement that lower summaries are not reliably estimable at that censoring
  level.

This is honest and still informative --- "arsenic was detected in 25% of samples;
the 90th percentile was 7.5 ug/L; the median was below the 2.0 ug/L reporting
limit" says exactly what the data support.

## Reporting imputed values

If you used `gsimp_impute()` to produce complete values (for a downstream model),
disclose it. State:

- **what** was imputed (which analytes/cells) and **why** (e.g. to fit a model
  that cannot accept intervals);
- the **method** --- a Gibbs sampler drawing each censored value from a truncated
  distribution within its interval, using the censored (`"tobit"`) conditional
  model --- and the settings (`imp_model`, `iters_all`, and, for multiple
  imputation, `M`);
- that multiple imputations were **pooled by Rubin's rules** (if you did so), and
  report the pooled estimate *with its multiple-imputation uncertainty*, not a
  single completed data set as if it were measured;
- a **reliability caveat**: imputation is dependable to roughly 50% non-detects
  (see `vignette("imputation-guidance")`); analytes beyond that were summarised
  descriptively instead.

For reproducibility, report the package version and the random seed:

```{r repro}
utils::packageVersion("leftcens")
```

### Citation

Cite the package and, because the imputation follows the GSimp method, the GSimp
paper:

```{r cite, eval = FALSE}
citation("leftcens")
```

> Wei R. *et al.* (2018) GSimp: A Gibbs sampler based left-censored missing value
> imputation approach for metabolomics studies. *PLoS Comput. Biol.*
> 14(1):e1005973.

## A reporting checklist

- Report **detection frequency** and the **MDL/LCMRL** for every analyte.
- Never substitute a number for a non-detect and then report ordinary
  statistics; report `< MDL` / `< LCMRL` or use the censoring-aware summaries.
- Report the **median** only where non-detects are below 50%; otherwise report it
  as `< LCMRL`.
- If you **imputed**, disclose the method, settings, and (for multiple
  imputation) Rubin pooling with its uncertainty; flag analytes beyond the
  reliable range as descriptive-only.
- Report the **package version**, the **random seed**, and cite the **GSimp
  paper** for the imputation.
