---
title: "Reporting censored-data results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reporting censored-data results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(leftcens)
```

This vignette is about **how to present** censored-measurement results in a report
or paper. It complements `vignette("imputation-guidance")`, which is about *when*
imputation is trustworthy; here we assume you have run the analysis and need to
write it up honestly.

## Non-detects: report the limit, not a substitute

The single most important rule: **do not replace non-detects with a made-up
number** (zero, the detection limit, or `MDL/2`) and then report ordinary
statistics on the result. That bias is exactly what `leftcens` exists to avoid.
Instead:

- Report a non-detect as **`< MDL`** (or as the interval), and state the method
  detection limit (MDL) and, where relevant, the quantitation limit (LCMRL).
- Report summary statistics computed by the censoring-aware tools
  (`desc_np()` / `desc_sp()`), not by substitution.
- If you imputed, say so explicitly (see below) --- imputed values are model
  outputs, not measurements.

## Which statistics to report, and when

Whether a statistic is even estimable depends on how much of the data is
censored, per analyte. A useful hierarchy:

- **Detection frequency** (the percentage detected) is *always* reportable and
  should be, whatever the censoring --- it is a fact about the data.
- **The median** is directly estimable when **fewer than 50% are non-detects**
  (the median then sits at or above the limit).
- **The mean and low quantiles** are trustworthy only under light censoring; the
  `desc_np()` NPMLE estimates them accounting for censoring, but they become
  unreliable as non-detects grow (see `vignette("imputation-guidance")`).

`leftcens` records the limits and returns quantiles below the quantitation limit
as `NA`, which is precisely what you want when tabulating: a blank/`< LCMRL`
rather than a spurious number. A reporting table falls out of `glance()` and
`quantile()`:

Report a *range* of quantiles, not just the median --- the low ones make the
censoring visible. Any quantile below the quantitation limit comes back as `NA`,
so the lower percentiles simply drop out for the more-censored analytes:

```{r report-table}
cl <- as_cens_list(groundwater)
np <- desc_np(cl)

g <- glance(np)
# include lower quantiles; those below the LCMRL come back as NA
q <- quantile(np, probs = c(0.1, 0.25, 0.5, 0.75, 0.9))

report <- data.frame(
  Analyte      = g$analyte,
  n            = g$nobs,
  `Detected %` = round(100 * (1 - g$n_non_detect / g$nobs)),
  LCMRL        = g$quantitation_limit,
  `10th`       = q[g$analyte, "10%"],
  `25th`       = q[g$analyte, "25%"],
  Median       = q[g$analyte, "50%"],
  `75th`       = q[g$analyte, "75%"],
  `90th`       = q[g$analyte, "90%"],
  check.names  = FALSE
)
knitr::kable(report, digits = 2,
             caption = "Only quantiles at or above the LCMRL carry a number; lower ones are NA (report as < LCMRL)")
```

Read every `NA` as "**< LCMRL**": that quantile falls below the quantitation
limit, so it is not a quantified number. Here the 10th and 25th percentiles are
`NA` for every metal, and the median is `NA` for several, while the upper
quantiles carry values. That *is* the honest summary: report the quantiles you
can, and `< LCMRL` for the rest. The **minimum** always lies below the detection
limit whenever any non-detects are present, so it is never a reportable number.

Note that **detected is not the same as quantified**: "detected" means above the
detection limit (MDL); "quantified" means above the quantitation limit (LCMRL).
An analyte can be mostly detected yet have its median in the
detected-not-quantified band --- arsenic here is 75% *detected* but its median is
still below the 2.0 LCMRL, hence `< LCMRL`. (A derived limit shown as `NA` simply
means that tier did not occur in the data --- e.g. an analyte with no non-detects
has no derivable MDL; report your laboratory's stated limit.)

## When an analyte is too censored for point summaries

If an analyte has more than roughly 50% non-detects, do **not** report an imputed
mean or median as if measured. Report instead:

- the **detection frequency** and the **MDL/LCMRL**;
- optionally an **interval-censored descriptive summary** (`desc_np()` gives the
  NPMLE of the distribution, and any estimable upper quantiles);
- a statement that lower summaries are not reliably estimable at that censoring
  level.

This is honest and still informative --- "arsenic was detected in 25% of samples;
the 90th percentile was 7.5 ug/L; the median was below the 2.0 ug/L reporting
limit" says exactly what the data support.

## Reporting imputed values

If you used `gsimp_impute()` to produce complete values (for a downstream model),
disclose it. State:

- **what** was imputed (which analytes/cells) and **why** (e.g. to fit a model
  that cannot accept intervals);
- the **method** --- a Gibbs sampler drawing each censored value from a truncated
  distribution within its interval, using the censored (`"tobit"`) conditional
  model --- and the settings (`imp_model`, `iters_all`, and, for multiple
  imputation, `M`);
- that multiple imputations were **pooled by Rubin's rules** (if you did so), and
  report the pooled estimate *with its multiple-imputation uncertainty*, not a
  single completed data set as if it were measured;
- a **reliability caveat**: imputation is dependable to roughly 50% non-detects
  (see `vignette("imputation-guidance")`); analytes beyond that were summarised
  descriptively instead.

For reproducibility, report the package version and the random seed:

```{r repro}
utils::packageVersion("leftcens")
```

### Citation

Cite the package and, because the imputation follows the GSimp method, the GSimp
paper:

```{r cite, eval = FALSE}
citation("leftcens")
```

> Wei R. *et al.* (2018) GSimp: A Gibbs sampler based left-censored missing value
> imputation approach for metabolomics studies. *PLoS Comput. Biol.*
> 14(1):e1005973.

## A reporting checklist

- Report **detection frequency** and the **MDL/LCMRL** for every analyte.
- Never substitute a number for a non-detect and then report ordinary
  statistics; report `< MDL` / `< LCMRL` or use the censoring-aware summaries.
- Report the **median** only where non-detects are below 50%; otherwise report it
  as `< LCMRL`.
- If you **imputed**, disclose the method, settings, and (for multiple
  imputation) Rubin pooling with its uncertainty; flag analytes beyond the
  reliable range as descriptive-only.
- Report the **package version**, the **random seed**, and cite the **GSimp
  paper** for the imputation.
