---
title: "Monte Carlo validation of `gsimp_impute()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5
)
results_file <- "results/latest.rds"
if (!file.exists(results_file)) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

```{r load, eval = file.exists("results/latest.rds")}
res_obj <- readRDS("results/latest.rds")
res <- res_obj$results
summ <- res_obj$summary
```

`r if (!file.exists(results_file)) "**No results found.** Run \`Rscript validation/mc_validation.R\` (or with a larger \`CONFIG=full N_REP=500\`) first, then knit this report."`

## Study design

This study evaluates how well `leftcens::gsimp_impute()` recovers known ground
truth for left- and interval-censored data, and whether its multiple-imputation
uncertainty is calibrated. Ground-truth log-concentrations are simulated
(multivariate, exchangeable correlation, optional sinh-arcsinh skew), censored
into the three-tier structure at analyte-specific limits, imputed, and compared
against the truth and against naive detection-limit substitution.

```{r config}
cat(sprintf(
  "config = %s | replications per scenario = %d | M (imputations) = %d | model = %s\n",
  res_obj$config, res_obj$n_rep, res_obj$M, res_obj$imp_model
))
```

## Headline: bias and MI coverage by scenario

```{r summary-table}
show <- summ[order(summ$censored_frac), c(
  "n", "p", "rho", "nd_frac", "skew", "censored_frac",
  "gs_bias", "gs_rmse", "naive_bias", "naive_rmse",
  "mi_coverage", "mi_coverage_mcse"
)]
show[] <- lapply(show, function(x) if (is.numeric(x)) round(x, 3) else x)
knitr::kable(show, row.names = FALSE,
             caption = "Per-scenario mean bias, RMSE (log scale), and 95% MI coverage of the true mean. Coverage should be ~0.95.")
```

## Point bias vs. censoring fraction

gsimp should reduce the upward bias that naive substitution introduces
(censoring removes the low values). The gap widens as censoring increases.

```{r bias-plot}
o <- order(summ$censored_frac)
cf <- summ$censored_frac[o]
plot(cf, summ$naive_bias[o], type = "b", pch = 16, col = "firebrick",
     ylim = range(0, summ$naive_bias, summ$gs_bias),
     xlab = "Censored fraction", ylab = "Mean bias (log scale)",
     main = "Recovered-mean bias: gsimp vs naive")
lines(cf, summ$gs_bias[o], type = "b", pch = 17, col = "steelblue")
abline(h = 0, lty = 3)
legend("topleft", c("naive (limit substitution)", "gsimp"),
       col = c("firebrick", "steelblue"), pch = c(16, 17), bty = "n")
```

## Calibration: multiple-imputation coverage

The key calibration check. A well-calibrated method's 95% MI interval covers the
true mean ~95% of the time. Watch where coverage degrades.

```{r coverage-plot}
plot(cf, summ$mi_coverage[o], type = "b", pch = 16, col = "darkgreen",
     ylim = c(0, 1),
     xlab = "Censored fraction", ylab = "95% MI coverage of the true mean",
     main = "MI coverage vs censoring fraction")
abline(h = 0.95, lty = 2, col = "grey40")
text(min(cf), 0.95, "nominal 0.95", pos = 3, cex = 0.8, col = "grey40")
```

## Correlation-structure recovery

```{r cor-plot}
plot(cf, summ$naive_cor_mae[o], type = "b", pch = 16, col = "firebrick",
     ylim = range(0, summ$naive_cor_mae, summ$gs_cor_mae, na.rm = TRUE),
     xlab = "Censored fraction", ylab = "Mean abs. error of correlations",
     main = "Between-analyte correlation recovery")
lines(cf, summ$gs_cor_mae[o], type = "b", pch = 17, col = "steelblue")
legend("topleft", c("naive", "gsimp"),
       col = c("firebrick", "steelblue"), pch = c(16, 17), bty = "n")
```

## Interpreting the results

- **Point bias.** gsimp consistently reduces the upward bias of naive
  substitution. Under light-to-moderate censoring the residual bias is small.
- **Heavy censoring is the limiting regime.** As the censored fraction grows,
  residual bias remains (the observed data simply carries less information about
  the low tail), and because that bias can exceed the multiple-imputation
  standard error, **MI coverage degrades** — the intervals are correctly narrow
  but centred slightly high. This is the honest boundary of the method and the
  main thing to report.
- **Levers to probe** (all supported by `mc_validation.R`): increasing
  `iters_all` (convergence), stronger inter-analyte correlation (more
  information for the conditional model), and the `imp_model` choice
  (`"glmnet"` for wide data). Re-running the grid across these isolates whether
  residual bias is a convergence artefact or an information limit.
```
